% Shakiba
% theory-application-diversity-measures.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% revision status as of 3/7/2026
% - [X] accept or reject all track changes
% - [X] incorporate review comments
%    - don't worry about clearing all comments
% - [X] incorporate reference suggestions
% - [X] cross-cutting revision objectives (see email)
% - [X] other planned changes or refinements (if any)
% for revision status, mark - for partial or x for complete
% any free-form comments on planned/completed revisions are helpful too!
% --------------------------------------------------------------------------------
% draft status: complete Draft as of <1/17/25>
% pick one of: outline, partial draft, rough draft, complete draft
% where rough draft indicates that major points are hit,
% but still working on changes or improvements in refining the text
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\section{Existing Connections: Diversity Measures}
\label{sec:diversity-measures}

% \textbf{Suggested Citations}: \begin{itemize}
% \item find full reference info in \texttt{filecontents} above, or in rendered PDF
% \item empty this list by moving citations into ``Included Citations'' or ``Excluded Citations'' below
% \end{itemize}
%
% \textbf{Included Citations}: \begin{itemize}
% \item move citations incorporated into section text here!
% \item \citet{burke2002advanced}
% \item \citet{burke2004diversity}
% \item \citet{morrison2002measurement}
% \item \citet{wineberg2003underlying}
% \item \citet{corriveau2012review}
% \item \citet{smit2011population}
% \item \citet{magurran2021measuring} --- very brief review worth skimming for other topics that may of interest in this section
% \end{itemize}
%
% \textbf{Excluded Citations}: \begin{itemize}
% \item move citations not incorporated into section text here!
% \item \citet{pugh2016quality}
% \item \citet{nguyen2006brief}
% \item \citet{mcphee1999analysis}
% \item \citet{gabor2018inheritancebased}
% \item \citet{rodriguezesparza2025addressing}
% \item \citet{mcphee2016visualizing,mcphee2016using}
% \item \citet{burlacu2023population}
% \item \citet{moreno2025ecology}
% \item \citet{kondratyeva2019reconciling}
% \item \citet{tian2019diversity} %Diversity assessment in MOO, not really biological
% \end{itemize}

In all branches of evolutionary computation, maintaining diversity in the population is a crucial factor for effectively solving problems because it prevents premature convergence to local optima and enables broader exploration of the search space.
Over the years, numerous strategies have been proposed to measure and maintain diversity within EC.
However, most existing research focuses only on the content of the population at a single time point \citep{hernandez2022can}.
These diversity measures are commonly referred to as ``genetic diversity'' or ``phenotypic diversity''.
In some cases, genetic diversity is quantified simply by counting the number of unique types present in the population at a given time point, a measure known in biology as richness \citep{magurran2021measuring}.
Other approaches employ metrics that account for how evenly individuals are distributed across types, such as Shannon diversity \citep{burke2002advanced, burke2004diversity}.
Occasionally more nuanced metrics are used that consider the level of similarity of the types in the population \citep{morrison2002measurement,corriveau2012review,smit2011population}.

In contrast, researchers in ecology and evolutionary biology augment these metrics with a class of phylogenetic diversity measures \citep{tucker2018relationship}, which account for not only individuals in the current population, but also their evolutionary history. %Phylogenetic diversity metrics quantify evolutionary divergence between individuals by incorporating information from the topology of the phylogenetic tree-a tree that represents the ancestral relationships among members of a population. Each node in the tree represents an individual, while the edges show ancestor-descendant relationships.
Similarly, in EC, by tracking the full phylogenetic tree of a population \citep{dolson2020interpreting}, analogous metrics can be derived to quantify how evolutionarily distant the candidate solutions are.
Existing work has explored two classes of phylogenetic diversity metrics in evolutionary computation:

\begin{itemize}
\item{Pairwise distance:} These metrics calculate the number of edges between a pair of nodes corresponding to the individuals in the population at a single time point \citep{webb2002phylogenies}.
\item{Evolutionary distinctiveness:} These metrics assign a score to each individual that reflects how evolutionarily unique it is compared to others \citep{isaac2007mammals}. The score is calculated by dividing the age of each branch in the phylogenetic tree by the number of nodes it leads to and then summing these values along the path from the individual to the root.
\end{itemize}

Previous research has shown that phylogenetic diversity metrics provide distinct information not captured by phenotypic diversity metrics.
In one study \citep{hernandez2022can}, five selection schemes were evaluated on several fitness landscapes.
First, the relationship between phenotypic richness (i.e., number of unique phenotypes in the final population) and mean pairwise distance was examined at the final generation.
The results showed a lack of consistent strong correlation between phenotypic and phylogenetic diversity at a fixed point in time.
This finding was further supported by analyzing how both types of diversity changed over time, showing that they exhibited different behaviors across longer temporal scales.
The same analysis was performed using different phenotypic and phylogenetic diversity metrics and similar results were observed.
%These results show that phylogenetic diversity provides novel information that phenotypic diversity does not.

Another experiment in the same study tested whether one type of diversity can be predictive of another.
This relationship was demonstrated using transfer entropy, a measure in information theory that captures the flow of information from one variable to another \citep{schreiber2000measuring}.
%Specifically, given two variables $X$ and $Y$, transfer entropy measures how much information past values of $Y$ provide about future values of $X$ beyond what is already explained by the past values of $X$.
Using this approach, it was shown that phylogenetic diversity can be a predictor of future phenotypic diversity.

Future phenotypic diversity is not the only variable that can be predicted using phylogenetic diversity.
Phylogenetic diversity has also been shown to be a predictor of future fitness.
Using the same transfer entropy approach, studies have demonstrated that in selection schemes that maintain diversity, phylogenetic diversity is more predictive of future success than phenotypic diversity \citep{hernandez2022can,shahbandegan2022untangling}.
This finding has important implications for evolutionary computation, as it suggests that monitoring phylogenetic diversity can provide early indicators of an algorithm's potential to discover global optima before it is run for thousands of generations.
Consequently, this can help choose the right evolutionary algorithm for an optimization problem, or enable the early identification of runs that are unlikely to be successful, reducing the need for computationally expensive long-term executions.

A rich ecosystem of existing software exists to support phylogenetic analyses and visualizations \citep{moreno2024dendropy,talevich2012biophylo,huertacepas2016ete3,aton2026scikitbio,kelleher2019inferring,moshiri2025compacttree,moshiri2020treeswift,ryneches2018suchtree,eaton2019toytree,moreno2026phyloframe,paradis2018ape,lalejini2019data,yu2016ggtree,zanini2025iplotx,sanderson2022taxonium,vaughan2017icytree}.
